An open reproducible framework for the study of the iterated prisoner's dilemma

### 2016-02-08 tex/paper.tex
Changing penultimate sections.
Merging the later two sections and various other clean ups.

Several iterated prisoner's dilemma tournaments have generated much interest,
including Axelrod's original tournaments, two 2004 anniversary tournaments, and
the Stewart and Plotkin 2012 tournament following the discovery of
zero-determinant strategies.  Subsequent research has spawned an enormous number
of papers, but rarely are the results reproducible. Among well-known
tournaments, in only one case is the full original source code available
(Axelrod's second tournament, in FORTRAN); in no cases is the available code
well-documented, easily modifiable, or released with significant test suites.

@@

This paper introduces a software package: the Axelrod python library
\cite{Axelrod-Pythonprojectteam2015}. The Axelrod-Python project has the
following stated goals:

@@

Before describing the package in more detail in
Section~\ref{sec:description-of-axelrod-python}, an overview of some previous
Iterated Prisoner's Dilemma research will be given.

@@

The game of (\ref{equ:one-shot}) is called the Prisoner's Dilemma. Numerical
values of \((R,S,T,P)=(3,0,5,1)\) are often used in the literature. Axelrod's
tournaments (and further implementations of these) are sometimes referred to as
Iterated Prisoner's Dilemma (IPD) tournaments. An overview of published tournaments is
given in Table~\ref{tab:tournaments}.

@@

\cite{Milgrom1982} describes how incomplete information can be used to
enhance cooperation, in a similar approach to the proof of the Folk theorem for
repeated games \cite{Maschler2013}. This aspect of incomplete information is
also considered in \cite{Molander1985, Bendor1991, Lee2015} where ``noisy''
tournaments randomly flip the choice made by a given strategy. In
\cite{Murnighan2015}, incomplete information is considered in the sense of a
probabilistic termination of each round of the tournament.

As mentioned before, IPD tournaments have been studied in an evolutionary
context: \cite{Ellison1994, Lee2015, Press2012, Stewart2012} consider this in a
traditional evolutionary game theory context. These works investigate
particular evolutionary contexts within which cooperation can emerge as stable.
Often these works consider direct opposition to another strategy and disregard
the population dynamics, for example in \cite{Lee2015} a simple machine learning
algorithm outperforms a strategy found in \cite{Press2012}.

Further to these evolutionary ideas, \cite{Chellapilla1999, DavidB1993} are
examples of using machine learning techniques to evolve particular strategies.
In \cite{Axelrod}, Axelrod describes how similar techniques are used to
genetically evolve a high performing strategy from a given set of strategies.
Note that in his original work, Axelrod only used a base strategy set of 12
strategies for this evolutionary study. This is noteworthy as
\cite{Axelrod-Pythonprojectteam2015} now boasts over 90 strategies that are
readily available for a similar analysis.

\subsection{Description of the Axelrod python package}\label{sec:description-of-axelrod-python}.

@@

Figure \ref{fig:demo_tournament_commands} just shows the very basic utilisation
of the library and further details can be found at the online documentation:
\url{http://axelrod.readthedocs.org}. Some further implemented capabilities
include:

\begin{itemize}
    \item Noisy tournaments
    \item Ecological tournaments
    \item Full tournament history
    \item Creation of plots of distributions of scores and wins
\end{itemize}

As stated in Section~\ref{sec:introduction} one of the main goals of the library
is to allow for the easy contribution of strategies. To do this requires the
writing of a simple python class (which can inherit from other predefined
classes). Full contribution guidelines can be found in the documentation.
Figures~\ref{fig:grudger} and~\ref{fig:grudger_test} show the source code for
the Grudger strategy as well as its corresponding test.

@@

You can see an overview of some of the source code in Figure~\ref{fig:overview}.

@@

To date the library has had contributions from 19 contributors from a variety of
backgrounds. These contributions have both been in terms of strategies (one
strategy is the creation of an undergraduate mathematics student with little
prior knowledge of programming) as well as the architecture of the library
itself.

Before discussing the novel insights obtained from the study of this library in
Section~\ref{sec:new-strategies-and-implications} an overview of some
tournaments that have been reproduced will be given in
Section~\ref{sec:reproducing-previous-tournaments}.

\section{Reproducing previous tournaments}\label{sec:reproducing-previous-tournaments}

@@

Work is ongoing to include more strategies. In the next Section, an
overview of some of the early and novel research insights made possible by the
library is described.

In parallel to the Python library, a tournament is being kept up to date that
pits all strategies against each other. Figure~\ref{fig:tournament} shows the
results from the full tournament.

@@

The current winning strategy is a novel strategy of type: \textbf{LookerUp}. This
is a strategy that maps a given set of states to actions. The state space is
defined generically by \(m, n\) so as to map states to actions in the following
way:

@@

The above example is an incomplete illustration of the mapping for \(m=8, n=2\).
Intuitively this state space uses the initial plays of the opponent to gain some
information about its intentions whilst still taking in to account the
recent play. The actual winning strategy is an instance of the framework for
\(m=n=2\) for which an evolutionary algorithm has been used to train it. More
details of this can be found in \cite{mojones}. In \cite{} experiments are
carried out to evaluate how this particular strategy behaves in environments
other than those in which it was trained\dots

@@

Details for the above can be found at \cite{}.

%Plot of memory_depth vs mean score, mean wins?

>>>>

Several iterated prisoner's dilemma tournaments have generated much interest,
including Axelrod's original tournaments, two 2004 anniversary tournaments, and
the Stewart and Plotkin 2012 tournament \cite{Stewart2012} following the
discovery of zero-determinant strategies.  Subsequent research has spawned an
enormous number of papers, but rarely are the results reproducible. Among
well-known tournaments, in only one case is the full original source code
available (Axelrod's second tournament, in FORTRAN); in no cases is the
available code well-documented, easily modifiable, or released with significant
test suites.

@@

This paper introduces a software package: the Axelrod Python library
\cite{Axelrod-Pythonprojectteam2015}. The Axelrod-Python project has the
following stated goals:

@@

Before describing the package in more detail in
Section~\ref{sec:description-of-axelrod-python}, an overview of some previous
Iterated Prisoner's Dilemma research will be given in
Section~\ref{sec:review}.

@@

The game of (\ref{equ:one-shot}) is called the Prisoner's Dilemma. Numerical
values of \((R,S,T,P)=(3,0,5,1)\) are often used in the literature. Axelrod's
tournaments (and further implementations of these) are sometimes referred to as
Iterated Prisoner's Dilemma (IPD) tournaments. An incomplete overview of
published tournaments is given in Table~\ref{tab:tournaments}.

@@

In \cite{Milgrom1982} a description is given of how incomplete information can
be used to enhance cooperation, in a similar approach to the proof of the Folk
theorem for repeated games \cite{Maschler2013}. This aspect of incomplete
information is also considered in \cite{Molander1985, Bendor1991, Lee2015} where
``noisy'' tournaments randomly flip the choice made by a given strategy. In
\cite{Murnighan2015}, incomplete information is considered in the sense of a
probabilistic termination of each round of the tournament.

As mentioned before, IPD tournaments have been studied in an evolutionary
context: \cite{Ellison1994, Lee2015, Press2012, Stewart2012} consider this in a
traditional evolutionary game theory context. These works investigate
particular evolutionary contexts within which cooperation can emerge as stable.
Often these works consider direct opposition to another strategy and disregard
the population dynamics, for example in \cite{Lee2015} a machine learning
algorithm outperforms a strategy found in \cite{Press2012}.

Further to these evolutionary ideas, \cite{Chellapilla1999, DavidB1993} are
examples of using machine learning techniques to evolve particular strategies.
In \cite{Axelrod}, Axelrod describes how similar techniques are used to
genetically evolve a high performing strategy from a given set of strategies.
Note that in his original work, Axelrod only used a base strategy set of 12
strategies for this evolutionary study. This is noteworthy as
\cite{Axelrod-Pythonprojectteam2015} now boasts over 120 strategies that are
readily available for a similar analysis.

\subsection{Description of the Axelrod Python package}\label{sec:description-of-axelrod-python}

@@

Figure~\ref{fig:demo_tournament_commands} just shows the very basic utilisation
of the library and further details can be found at the online documentation:
\url{http://axelrod.readthedocs.org}.

As stated in Section~\ref{sec:introduction} one of the main goals of the library
is to allow for the easy contribution of strategies. To do this requires the
writing of a simple Python class (which can inherit from other predefined
classes). Full contribution guidelines can be found in the documentation.
As an example, Figures~\ref{fig:grudger} and~\ref{fig:grudger_test} show the source code for
the Grudger strategy as well as its corresponding test.
You can see an overview of the structure of the source code in Figure~\ref{fig:overview}.

@@

@@

To date the library has had contributions from 24 contributors from a variety of
backgrounds. These contributions have both been in terms of strategies (one
strategy is the creation of an undergraduate mathematics student with little
prior knowledge of programming) as well as the architecture of the library
itself.

Section~\ref{sec:new-strategies-and-implications} will describe in more detail
some of research capabilities of the library.

@@

In parallel to the Python library, a tournament is being kept up to date that
pits all available strategies against each other. Figure~\ref{fig:tournament}
shows the results from the full tournament which can also be seen (in full
detail) here: \url{http://axelrod-tournament.readthedocs.org/}.

@@

The current winning strategy is a novel strategy called: Looker Up. This
is a strategy that maps a given set of states to actions. The state space is
defined generically by \(m, n\) so as to map states to actions in the following
way:

@@

The above example is an incomplete illustration of the mapping for \(m=8, n=2\).
Intuitively, this state space uses the initial plays of the opponent to gain
some information about its intentions whilst still taking in to account the
recent play. The actual winning strategy is an instance of the framework for
\(m=n=2\) for which an evolutionary algorithm has been used to train it. More
details of this can be found at
\url{http://mojones.net/evolving-strategies-for-an-iterated-prisoners-dilemma-tournament.html}.
At
\url{http://vknight.org/unpeudemath/gametheory/2015/11/28/Experimenting-with-a-high-performing-evolved-strategy-in-other-environments/}
experiments are described that evaluate how this particular strategy behaves in
environments other than those in which it was trained\dots

@@

Details for the above can be found at \url{http://marcharper.codes/}.
